---
title: "Final Project Proposal"
author: "Ava, Alex, Luci, and Christine"
date: "5-31-26!"
output: pdf_document
---


#check to make sure say per 100,000 everywhere instead of per capita 

## Final Project Proposal
Proposals should answer the following questions:

1. What is your research question? Why does it matter? How does it relate to society?

2. What data will you use to answer the question? Do you have access to these data?

3. What methods will you use to answer your research question?

4. What kind of product will you create? How do you want others to read or engage with your project?


Download all necessary libraries
```{r libraries, echo = T, message=F, warning=F}
#Libraries
library(RCurl)
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(sf)
library(tidycensus)
library(tigris)
library(viridis)

```


Download data 
```{r data, echo = T, message=F, warning=F}

url <- getURL("https://raw.githubusercontent.com/avaelisa/soc10finalproject-alex-christine-ava-luci/refs/heads/main/soc10-dataset")

# download incarceration data from the BJS (from github)
data <- read_csv(url)

head(data) #uncleaned data downloaded from the BJS

```

Data processing part 1: fix names, prep per 100,000 people, and spatial data 

```{r data processing 1, echo = T, message=F, warning=F}

#Mutate to fix state names (remove data type signifiers) and remove U.S. total, Federal, and All States from BJS dataset
data %<>%
  mutate(Location=case_match(Location,
                             "Alaska/b"~"Alaska",
                             "Connecticut/b"~"Connecticut",
                             "Delaware/b"~"Delaware",
                             "Hawaii/b"~"Hawaii",
                             "Maryland/c"~"Maryland",
                             "Missouri/d"~"Missouri",
                             "Oklahoma/c"~"Oklahoma",
                             "Rhode Island/b"~"Rhode Island",
                             "Vermont/b"~"Vermont",
                             .default=Location)) %>%
                            drop_na(Total) %>% #get rid of district of columbia and puerto rico because there is no data
  filter(!Location %in% c("U.S. total", "Federal", "All States"))

head(data) #cleaned data


#census API
census_api_key("857ead205d4b6101ae723a246e9b6f95b1b827d4")

#add per 100,000 with tidycensus
states <- get_acs(
  geography = "state",
  variables = "B01003_001",
  geometry = TRUE,
  year = 2023) %>%
  filter(!NAME %in% c("District of Columbia", "Puerto Rico")) #remove these because no data as seen above 

head(states) #tidy census state data from 2023 (matching 2023 BJS data)

#combine data with left_join USA map + incarceration rate
states_incar <- states %>%
  left_join(data, by = c("NAME" = "Location"))

head(states_incar) #combined tidycensus and BJS data 

#add female incarceration numbers per one hundred thousand people to the spatial data
states_incar_fpc <- states_incar %>%
  mutate("Female Incarceration Ratio Per 100,000" = Female/estimate*100000)

head(states_incar_fpc) #new data set includes female incarceration ration per 100,000 in state population

#create column of incarceration per 1000,000 males 
#add male incarceration numbers per one hundered thousand people to the spatial data
states_incar_fmpc <- states_incar_fpc %>%
  mutate("Male Incarceration Ratio Per 100,000" = Male/estimate*100000)

head(states_incar_fmpc) #new data set includes male incarceration ratio per 100,000 in state population

```


Bar Chart 1A: Total Population Incarcerated by State
Goal: Compare incarceration rates for all states

Using coord_flip to make bar chart horizontal. Reorder in ggplot will rank bars from smallest to biggest so it is easier to see trends and which states have highest vs lowest incarceration rates. This will make data easier to analyze. 

```{r figure 1A, echo = T, message=F, warning=F}

state_total_graph <- ggplot(data, aes(x = reorder(Location, Total), y = Total))+ #reorder 
  geom_bar(position = "dodge", stat = "identity", 
           aes(fill = Location))+
           labs(y = "Total", x="State", title = "Total Population Incarcerated by State")+
             coord_flip() 


#See visualization 
state_total_graph 

```


Bar Chart 1B: Number of Females Incarcerated Per 100,000 People in the State
Goal: Compare female incarceration ratios for all states

```{r figure 1B, echo = T, message=F, warning=F}

state_female_ratio_graph <- ggplot(states_incar_fpc, aes(x = reorder(NAME, `Female Incarceration Ratio Per 100,000`), y = `Female Incarceration Ratio Per 100,000`))+ #reorder 
  geom_bar(position = "dodge", stat = "identity", 
           aes(fill = NAME))+
           labs(x = "State", y = "Total", title = "Number of Females Incarcerated Per 100,000 People in the State")+
             coord_flip() 


#See visualization 
state_female_ratio_graph

```



Bar Chart 2A: Total Number of People Incarcerated for Each Region 

Goal: Divide state data into their respective regions (Northeast, Midwest, West, South) and add number of people incarcerated for each region

The states were divided into regions through case_match. Each state's incarceration numbers for each region were summed to calculate total incarceration numbers for each region.  

```{r figure 2, echo = T, message=F, warning=F}

region_data <- data %>% 
  mutate(Region=case_match(Location,
                             #Northeast
                             c("Maine", "Vermont", "New Hampshire", "Massachusetts", "Connecticut", "Rhode Island", "New York", "Pennsylvania", "New Jersey", "Virginia") ~ "Northeast",
                             #Midwest
                             c("North Dakota", "South Dakota", "Nebraska", "Kansas", "Minnesota", "Iowa", "Missouri", "Illinois", "Wisconsin", "Michigan", "Ohio", "Indiana") ~ "Midwest",
                             #West
                             c("Washington", "Oregon", "California", "Nevada", "Idaho", "Wyoming", "Montana", "Colorado", "Utah", "Arizona", "New Mexico") ~ "West",
                             #South
                             c("Texas", "Oklahoma", "Arkansas", "Louisiana", "Mississippi", "Kentucky", "Tennessee", "Alabama", "Georgia", "West Virginia", "Florida", "South Carolina", "North Carolina", "Maryland", "Delaware") ~ "South",
                             .default = "Other")) %>%
                            drop_na(Total) %>%
  group_by(Region) %>% mutate(sum = sum(Total, na.rm = TRUE))

#View(region_data)

head(region_data) #states' data are now organized into their respective regions

#Plot
region_total_graph <- 
  ggplot(region_data, aes(x = reorder(Region, Total), y = sum)) +
  geom_bar(aes(fill = Region), position = "dodge", stat = "identity")+
  labs(x = "Region", y = "Total", title = "Total Number of People Incarcerated by Region")+
  theme_linedraw()+
  coord_flip()

#See visualization
region_total_graph

```



Bar Chart 2B: Number of Females Incarcerated Per 100,000 People in Population for Each Region 

Goal: To show the ratio of females incarcerated per 100,000 people in each region. This is important because this data takes into consideration total population size.  

```{r figure 2B, echo = T, message=F, warning=F}


region_data_pc <- states_incar_fmpc %>% 
  mutate(Region=case_match(NAME,
                             #Northeast
                             c("Maine", "Vermont", "New Hampshire", "Massachusetts", "Connecticut", "Rhode Island", "New York", "Pennsylvania", "New Jersey", "Virginia") ~ "Northeast",
                             #Midwest
                             c("North Dakota", "South Dakota", "Nebraska", "Kansas", "Minnesota", "Iowa", "Missouri", "Illinois", "Wisconsin", "Michigan", "Ohio", "Indiana") ~ "Midwest",
                             #West
                             c("Washington", "Oregon", "California", "Nevada", "Idaho", "Wyoming", "Montana", "Colorado", "Utah", "Arizona", "New Mexico") ~ "West",
                             #South
                             c("Texas", "Oklahoma", "Arkansas", "Louisiana", "Mississippi", "Kentucky", "Tennessee", "Alabama", "Georgia", "West Virginia", "Florida", "South Carolina", "North Carolina", "Maryland", "Delaware") ~ "South",
                             .default = "Other")) %>%
                            drop_na(Total) %>%
  group_by(Region) %>% mutate(sum = sum(`Female Incarceration Ratio Per 100,000`, na.rm = TRUE))

#View(region_data_pc)

head(region_data_pc) #Ratio of females incarcerated per 100,000 people by region

#Plot
region_total_graph <- 
  ggplot(region_data_pc, aes(x = reorder(Region, estimate), y = sum)) +
  geom_bar(aes(fill = Region), position = "dodge", stat = "identity")+
  labs(x = "Region", y = "Total", title = "Number of Females Incarcerated Per 100,000 People in Population for Each Region ")+
  theme_linedraw()+
  coord_flip()

#Show visualization
region_total_graph

```

Bar Chart 3A: Total Incarceration Numbers with States Divided by State by Sex

```{r figure 3A, echo = T, message=F, warning=F}

#pivot longer
state_data <- data %>%
  pivot_longer(cols = c(Male, Female),
               names_to = "Sex",
               values_to = "Value")
 
#plot              
state_total_stacked <- ggplot(state_data, aes(x = Location, y = Value))+
  geom_bar(position = "stack", stat = "identity",
           aes(fill = Sex))+
  theme_linedraw()+
  labs(x = "State", title = "Total Population Incarcerated in State by Sex")+
  coord_flip() 
  
#Show visualization
state_total_stacked

```

Bar Chart 3B: Number of Males and Females Incarcerated Per 100,000 People by State
Goal: 

```{r figure 3B, echo = T, message=F, warning=F}

#pivot longer
gendered_states_incar_fmpc <- states_incar_fmpc %>%
  pivot_longer(cols = c(`Male Incarceration Ratio Per 100,000`, `Female Incarceration Ratio Per 100,000`),
               names_to = "Sex",
               values_to = "Value")
 
#plot              
state_total_stacked <- ggplot(gendered_states_incar_fmpc, aes(x = NAME, y = Value))+
  geom_bar(position = "stack", stat = "identity",
           aes(fill = Sex))+
  theme_linedraw()+
  labs(x = "State", title = "Number of Males and Females Incarcerated Per 100,000 People by State and Sex")+
  coord_flip() 
  
#Show visualization
state_total_stacked

```




Bar Chart 4A: Total Incarceration Rate with States Divided by Region by Sex

```{r figure 4A, echo = T, message=F, warning= F}

#pivot longer
region_data2 <- region_data %>%
  pivot_longer(cols = c(Male, Female),
               names_to = "Sex",
               values_to = "Value")

#plot 
sex_by_region <- ggplot(region_data2, aes(x = Region, y = Value))+
  geom_bar(position = "stack", stat = "identity",
           aes(fill = Sex))+
  theme_linedraw()+
  labs(x = "Region", title = "Total Population Incarcerated by Region and Sex")+
  coord_flip() 
 
#print plot  
sex_by_region

```
Bar Chart 4B: Number of Males and Females Incarcerated Per 100,000 People in Each Region

```{r figure 4B, echo = T, message=F, warning= F}

#pivot longer
region_data_pc2 <- region_data_pc %>%
  pivot_longer(cols = c(`Male Incarceration Ratio Per 100,000`, `Female Incarceration Ratio Per 100,000`),
               names_to = "Sex",
               values_to = "Value")

#plot 
sex_by_region_pc_graph <- ggplot(region_data_pc2, aes(x = Region, y = Value))+
  geom_bar(position = "stack", stat = "identity",
           aes(fill = Sex))+
  theme_linedraw()+
  labs(x = "Region",  title = "Total Number of Males and Females Incarcerated in each Region Per 100,000 People")+
  coord_flip() 
 
#See Visualization  
sex_by_region_pc_graph

```


Spatial Graph 1: Number of Females Incarcerated per 100,000 People in Each State

```{r figure 5, echo = T, message=F, warning=F}

#plot combined data - Female Incarceration Ratio Per 100,000 People in Each State
ggplot(states_incar_fpc)+
  geom_sf(aes(fill = `Number of Females Incarcerated Per 100,000 People in Each State`))+
  scale_fill_viridis(option = "magma", direction = -1)+
  labs(title = "Female Incarceration Ratio Per 100,000 in Each State") +
  coord_sf(xlim = c(-185, -60))

```


Spatial Graph 2: Number of Females Incarcerated to Number of Males Incarcerated per 100,000 People in Each State

```{r figure 6, echo = T, message=F, warning=F}

#add female per male to spatial data 

states_incar_fpm <- states_incar_fmpc %>%
  mutate("Female per 100,000 population to Male per 100,000 Ratio" = `Female Incarceration Ratio Per 100,000`/ `Male Incarceration Ratio Per 100,000`)

head(states_incar_fpm)

#plot combined data - Ratio of Female to Male
ggplot(states_incar_fpm)+
  geom_sf(aes(fill = `Female per 100,000 population to Male per 100,000 Ratio`))+
  scale_fill_viridis(option = "magma", direction = -1)+
  labs(title = "Ratio of Females to Males Incarcerated in Each State")  +
  coord_sf(xlim = c(-185, -60))

```

  

Spatial Graph 3: Highlighting states with specific policies on spatial graph of total # of female incarcerated to per 100,000 people's arrests

```{r figure 7, echo = T, message=F, warning=F}

#highlight specific states 

strict_prositution_states <- states_incar_fpc %>% 
  filter(NAME %in% c("Oklahoma", "Texas", "Utah", "Pennsylvaia", "Nevada", "Kentucky", "West Virginia", "Vermont", "Louisiana", "Idaho", "Indiana", "Illinois"))

head(strict_prositution_states)

#Number of Females Incarcerated Per 100,000 People in Each State, Highlighting States with Strict Prostitution Laws in Green
ggplot(states_incar_fpc)+
  geom_sf(aes(fill = `Female Incarceration Ratio Per 100,000`))+
  scale_fill_viridis(option = "magma", direction = -1)+
  labs(title = "Number of Females Incarcerated Per 100,000 People in Each State", 
       subtitle = "States with Stricter Stances on Prostitution Laws are Highlighted in Green Borders")+
  geom_sf(data = strict_prositution_states, 
          fill = NA, 
          color = "green",
          size = 0.5) +
  coord_sf(xlim = c(-185, -60))

```

  
T-Test: Highlighting states with specific policies on spatial graph of total # of female incarcerated to male incarcerated per state

p-value is typically less than 0.05 (or 5%), the result is considered statistically significant. used default two tailed t-test. 

```{r t-test, echo = T, message=F, warning=F}

#add row for those with stricter take on prostitution vs not
strictness_states_incar_fpc <- states_incar_fpc %>% 
  mutate(Strictness=case_match(NAME,
                             #Strict
                             c("Oklahoma", "Texas", "Utah", "Pennsylvaia", "Nevada", "Kentucky", "West Virginia", "Vermont", "Louisiana", "Idaho", "Indiana", "Illinois") ~ "Strict",
                             c("South Carolina", "Maine", "New Hampshire", "Massachusetts", "Connecticut", "Rhode Island", "New York", "California", "New Jersey", "North Dakota", "South Dakota", "Nebraska", "Kansas", "Minnesota", "Iowa", "Missouri", "Illinois", "Wisconsin", "Michigan", "Ohio", "Washington", "Oregon", "Wyoming", "Montana", "Colorado", "Arizona", "New Mexico", "Arkansas", "Mississippi", "Tennessee", "Alabama", "Georgia", "Florida", "North Carolina", "Maryland", "Delaware", "Pennsylvania", "Alaska", "Hawaii", "Virginia") ~ "Not Strict", 
         .default = "Other")) %>%
  group_by(Strictness) %>% mutate(sum = sum(Total, na.rm = TRUE))

head(strictness_states_incar_fpc)

#ttest 
t.test(`Female Incarceration Ratio Per 100,000` ~ Strictness, data = strictness_states_incar_fpc)


```

