---
title: "Final Project Proposal"
author: "Ava, Alex, Luci, and Christine"
date: "5-31-26!"
output: pdf_document
---

## Final Project Proposal
Proposals should answer the following questions:

1. What is your research question? Why does it matter? How does it relate to society?

2. What data will you use to answer the question? Do you have access to these data?

3. What methods will you use to answer your research question?

4. What kind of product will you create? How do you want others to read or engage with your project?


Download all necessary libraries
```{r libraries, echo = T, message=F, warning=F}
#Libraries
library(RCurl)
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(sf)
library(tidycensus)
library(tigris)
library(viridis)
library(magrittr)

```


Download data 
```{r data, echo = T, message=F, warning=F}

url <- getURL("https://raw.githubusercontent.com/avaelisa/soc10finalproject-alex-christine-ava-luci/refs/heads/main/soc10-dataset")

# download incarceration data from the BJS (from github)
data <- read_csv(url)

head(data) #uncleaned data downloaded from the BJS

```

Data processing part 1: fix names, prep per 100,000 people, and spatial data 

```{r data processing 1, echo = T, message=F, warning=F}

#Mutate to fix state names (remove data type signifiers) and remove U.S. total, Federal, and All 
#States from BJS dataset
data %<>%
  mutate(Location=case_match(Location,
                             "Alaska/b"~"Alaska",
                             "Connecticut/b"~"Connecticut",
                             "Delaware/b"~"Delaware",
                             "Hawaii/b"~"Hawaii",
                             "Maryland/c"~"Maryland",
                             "Missouri/d"~"Missouri",
                             "Oklahoma/c"~"Oklahoma",
                             "Rhode Island/b"~"Rhode Island",
                             "Vermont/b"~"Vermont",
                             .default=Location)) %>%
                            drop_na(Total) %>% #Get rid of Washington D.C. and Puerto Rico 
  filter(!Location %in% c("U.S. total", "Federal", "All States"))

head(data) #Cleaned data


#Census API
census_api_key("857ead205d4b6101ae723a246e9b6f95b1b827d4")

#Add spatial data and estimate data with tidycensus
states <- get_acs(
  geography = "state",
  variables = "B01003_001",
  geometry = TRUE,
  year = 2023) %>%
  filter(!NAME %in% c("District of Columbia", "Puerto Rico")) #remove these here as well

head(states) #Tidycensus state data from 2023 (matching 2023 BJS data)

#Combine data with left_join USA map + incarceration rate
states_incar <- states %>%
  left_join(data, by = c("NAME" = "Location"))

head(states_incar) #Combined tidycensus and BJS data 

#add number of females incarcerated per one hundred thousand people to the spatial data
states_incar_fpc <- states_incar %>%
  mutate("Number of Females Incarcerated Per 100,000 People" = Female/(estimate*100000))

head(states_incar_fpc) #New data set includes Number of Females Incarcerated Per 100,000 
#People in Each State in state population

#Create column of number of males incarcerated per 1000,000 people into dataset with 
#spatial data and Number of Females Incarcerated Per 100,000 People in Each State
states_incar_fmpc <- states_incar_fpc %>%
  mutate("Number of Males Incarcerated Per 100,000 People" = Male/(estimate*100000))

head(states_incar_fmpc) #New data set includes number males incarcerated per 100,000 
#people in the state population

#total number of people incarcerated per population in each state
states_incar_ipp <- states_incar %>%
  mutate("Number of People Incarcerated Per 100,000 People" = Total/(estimate*100000))

#preparing region data 
states_incar_region <- states_incar %>% 
  mutate(Region=case_match(NAME,
                             #Northeast
                             c("Maine", "Vermont", "New Hampshire", "Massachusetts", "Connecticut", 
                               "Rhode Island", "New York", "Pennsylvania", "New Jersey") 
                           ~ "Northeast",
                             #Midwest
                             c("North Dakota", "South Dakota", "Nebraska", "Kansas", "Minnesota", 
                               "Iowa", "Missouri", "Illinois", "Wisconsin", "Michigan", "Ohio", 
                               "Indiana") ~ "Midwest",
                             #West
                             c("Washington", "Oregon", "California", "Nevada", "Idaho", "Wyoming", 
                               "Montana", "Colorado", "Utah", "Arizona", "New Mexico") ~ "West",
                             #South
                             c("Texas", "Oklahoma", "Arkansas", "Louisiana", "Mississippi", 
                               "Kentucky", "Tennessee", "Alabama", "Georgia", "West Virginia", 
                               "Florida", "South Carolina", "North Carolina", "Maryland", 
                               "Delaware","Virginia") ~ "South",
                             .default = "Other")) %>% #alaska and hawaii
                            drop_na(Total) %>%
  group_by(Region) %>% mutate(incarsum = sum(Total, na.rm = TRUE)) %>%
  group_by(Region) %>% mutate(popsum = sum(estimate, na.rm = TRUE)) %>%
  group_by(Region) %>% mutate(fincarsum = sum(Female, na.rm = TRUE)) %>%
  group_by(Region) %>% mutate(mincarsum = sum(Male, na.rm = TRUE)) 

states_incar_region2 <- states_incar_region %>%
 mutate("Number of Females Incarcerated Per 100,000 People" = fincarsum/(popsum*100000))%>%
 mutate("Number of Males Incarcerated Per 100,000 People" = mincarsum/(popsum*100000)) %>%
 mutate("Number of People Incarcerated Per 100,000 People" = incarsum/(popsum*100000))
  
head(states_incar_region2) #States' data are now organized into their respective regions

```


Bar Chart 1A: Total Population Incarcerated by State
Goal: Compare incarceration totals across all states

Using coord_flip to make bar chart horizontal. Reorder in ggplot will rank bars from smallest to biggest so it is easier to see trends and which states have highest vs lowest incarceration rates. This will make data easier to analyze. 

```{r figure 1A, echo = T, message=F, warning=F}

state_total_graph <- ggplot(data, aes(x = reorder(Location, Total), y = Total))+ #Reorder 
  geom_bar(position = "dodge", stat = "identity")+
           labs(y = "Total", x="State", title = "Total Population Incarcerated by State")+
             coord_flip() +
  theme(legend.position = "none")


#See visualization 
state_total_graph 

```
Bar Chart 1B: Number of People Incarcerated Per 100,000 People in the State
Goal: Compare total incarceration according to population for all states

```{r figure 1b, echo = T, message=F, warning=F }
        
state_pop_ratio_graph <- ggplot(states_incar_ipp, 
                                aes(x = reorder(NAME, `Number of People Incarcerated Per 100,000 People`), 
                                    y = `Number of People Incarcerated Per 100,000 People`))+ #reorder 
  geom_bar(position = "dodge", stat = "identity")+
           labs(x = "State", y = "Total", 
                title = "Number of People Incarcerated/100,000 People in Each State")+
             coord_flip() +
  theme(legend.position = "none")


#See visualization 
state_pop_ratio_graph


```

Bar Chart 1C: Number of Females Incarcerated Per 100,000 People in the State
Goal: Compare female incarceration according to population for all states

```{r figure 1C, echo = T, message=F, warning=F}

state_female_ratio_graph <- ggplot(states_incar_fpc, 
    aes(x = reorder(NAME, `Number of Females Incarcerated Per 100,000 People`), 
        y = `Number of Females Incarcerated Per 100,000 People`))+ #reorder 
  geom_bar(position = "dodge", stat = "identity") +
           labs(x = "State", y = "Total", 
                title = "Number of Females Incarcerated Per 100,000 People in Each State")+
             coord_flip() +
  theme(legend.position = "none")


#See visualization 
state_female_ratio_graph

```

Bar Chart 2A: Total Number of People Incarcerated for Each Region (and preparing region data)

Goal: Divide state data into their respective regions (Northeast, Midwest, West, South) and add number of people incarcerated for each region

The states were divided into regions through case_match. Each state's incarceration numbers for each region were summed to calculate total incarceration numbers for each region.  

```{r figure 2, echo = T, message=F, warning=F}

#Plot
region_total_graph <- 
  ggplot(states_incar_region2, aes(x = reorder(Region, incarsum), y = incarsum)) +
  geom_bar(aes(fill = Region), position = "dodge", stat = "identity")+
  labs(x = "Region", y = "Total", title = "Total Number of People Incarcerated by Region")+
  theme_linedraw()+
  coord_flip()

#See visualization
region_total_graph

```


Bar Chart 2B: Number of People Incarcerated Per 100,000 People in Population for Each Region 

Goal: To show the Number of people incarcerated per 100,000 people in each region. This is important because this data takes into consideration total population size.  

```{r figure 2B, echo = T, message=F, warning=F}

#Plot
region_ipp_graph <- 
  ggplot(states_incar_region2, 
         aes(x = reorder(Region, `Number of People Incarcerated Per 100,000 People`), 
             y = `Number of People Incarcerated Per 100,000 People`)) +
  geom_bar(aes(fill = Region), position = "dodge", stat = "identity")+
  labs(x = "Region", y = "Number of People Incarcerated Per 100,000 People in Each Region", 
       title = "Number of People Incarcerated Per 100,000 People in Each Region")+
  theme_linedraw()+
  coord_flip()

#Show visualization
region_ipp_graph

```


Bar Chart 2C: Number of Females Incarcerated Per 100,000 People in Population for Each Region 

Goal: To show the Number of females incarcerated per 100,000 people in each region. This is important because this data takes into consideration total population size.  

```{r figure 2C, echo = T, message=F, warning=F}

#Plot
region_totals_graph <- 
  ggplot(states_incar_region2, 
         aes(x = reorder(Region, `Number of Females Incarcerated Per 100,000 People`), 
             y = `Number of Females Incarcerated Per 100,000 People`)) +
  geom_bar(aes(fill = Region), position = "dodge", stat = "identity")+
  labs(x = "Region", y = "Number of Females Incarcerated Per 100,000 People in Each Region", 
       title = "Number of Females Incarcerated Per 100,000 People in Each Region")+
  theme_linedraw()+
  coord_flip()

#Show visualization
region_totals_graph

```

Bar Chart 3A: Total Population Incarcerated in Each State by Sex
Goal: To show sex differences in total incarcerated population in each state

```{r figure 3A, echo = T, message=F, warning=F}

#Pivot longer
state_data <- data %>%
  pivot_longer(cols = c(Male, Female),
               names_to = "Sex",
               values_to = "Value")
 
#plot              
state_total_stacked <- ggplot(state_data, aes(x = Location, y = Value))+
  geom_bar(position = "stack", stat = "identity",
           aes(fill = Sex))+
  theme_linedraw()+
  labs(x = "State", y = "Total Population Incarcerated", 
       title = "Total Population Incarcerated in Each State by Sex")+
  coord_flip() 
  
#Show visualization
state_total_stacked

```

Bar Chart 3B: Number of People Incarcerated Per 100,000 People in the State's Population by Sex
Goal: To show the Number of females incarcerated per 100,000 people in each state by sex. This is important because this data takes into consideration total population size.  

```{r figure 3B, echo = T, message=F, warning=F}

#Pivot longer
gendered_states_incar_fmpc <- states_incar_fmpc %>%
  pivot_longer(cols = c(`Number of Males Incarcerated Per 100,000 People`, 
                        `Number of Females Incarcerated Per 100,000 People`),
               names_to = "Sex",
               values_to = "Value") %>% mutate(Sex=
                case_match(Sex, "Number of Males Incarcerated Per 100,000 People" ~ "# of Male/100,000", 
                      "Number of Females Incarcerated Per 100,000 People" ~ "# of Female/100,000"))
 
#Plot              
state_total_stacked <- ggplot(gendered_states_incar_fmpc, aes(x = NAME, y = Value))+
  geom_bar(position = "stack", stat = "identity",
           aes(fill = Sex))+
  theme_linedraw()+
  labs(x = "State", y = "Number of People Incarcerated Per 100,000 People in the State's Population", 
       title = "Number of People Incarcerated/100,000 People in the State's Pop by Sex")+
  coord_flip() +
  theme(legend.position = "bottom")
  
#Show visualization
state_total_stacked

```




Bar Chart 4A: Total Population Incarcerated in Each Region by Sex
Goal: To show sex differences in total incarcerated population in each region

```{r figure 4A, echo = T, message=F, warning= F}

#Pivot longer
states_incar_region3 <- states_incar_region2 %>%
  pivot_longer(cols = c(Male, Female),
               names_to = "Sex",
               values_to = "Value")

#Plot 
sex_by_region <- ggplot(states_incar_region3, aes(x = Region, y = Value))+
  geom_bar(position = "stack", stat = "identity",
           aes(fill = Sex))+
  theme_linedraw()+
  labs(x = "Region", y= "Total Population Incarcerated in Each Region", 
       title = "Total Population Incarcerated in Each Region by Sex")+
  coord_flip() 
 
#Print plot  
sex_by_region

```
Bar Chart 4B: Total Number of People Incarcerated Per 100,000 People in Each Region by Sex
To show the Number of females incarcerated per 100,000 people in each region by sex. This is important because this data takes into consideration total population size.  

```{r figure 4B, echo=T, message=FALSE, warning=FALSE}

#Pivot longer
region_data_fpc <- states_incar_region2 %>%
  pivot_longer(cols = c(`Number of Females Incarcerated Per 100,000 People`, 
                        `Number of Males Incarcerated Per 100,000 People`),
               names_to = "Sex",
               values_to = "Value") %>% mutate(Sex=
                case_match(Sex, "Number of Males Incarcerated Per 100,000 People" ~ "# of Male/100,000", 
                      "Number of Females Incarcerated Per 100,000 People" ~ "# of Female/100,000"))

#Plot 
sex_by_region_pc_graph <- ggplot(region_data_fpc, aes(x = Region, y = Value))+
  geom_bar(position = "stack", stat = "identity",
           aes(fill = Sex))+
  theme_linedraw()+
  labs(x = "Region", y= "Number of People Incarcerated/100,000 people", 
       title = "Total Number of People Incarcerated/100,000 People in Each Region by Sex")+
  coord_flip() +
  theme(legend.position = "bottom")
 
#See Visualization  
sex_by_region_pc_graph

```


Spatial Graph 1: Number of Females Incarcerated per 100,000 People in Each State

```{r figure 5, echo = T, message=F, warning=F}

#Plot spatial and Number of Females Incarcerated Per 100,000 People in Each State data
ggplot(states_incar_fpc)+
  geom_sf(aes(fill = `Number of Females Incarcerated Per 100,000 People`))+
  scale_fill_viridis(option = "magma", direction = -1)+
  labs(title = "Number of Females Incarcerated Per 100,000 People in Each State") +
  coord_sf(xlim = c(-185, -60)) +
  theme(legend.position = "bottom")

```


Spatial Graph 2: Number of Females Incarcerated per 100,000 people in each state to Number of Males Incarcerated per 100,000 People in Each State

```{r figure 6, echo = T, message=F, warning=F}

#Add Number of Females Incarcerated per 100,000 in State Population to 
#Number of Males Incarcerated per 100,000 in State Population

states_incar_fpm <- states_incar_fmpc %>%
  mutate("Female per 100,000 population to Male per 100,000" = 
           `Number of Females Incarcerated Per 100,000 People`/ 
           `Number of Males Incarcerated Per 100,000 People`)

head(states_incar_fpm) #New data set with # of females incarcerated 
#in a state /# of males incarcerated in a state

#plot combined data - Female to Male
ggplot(states_incar_fpm)+
  geom_sf(aes(fill = `Female per 100,000 population to Male per 100,000`))+
  scale_fill_viridis(option = "magma", direction = -1)+
  labs(title = "Number of Females Incarcerated/100,000 people per
Number of Males Incarcerated/100,000 people")  +
  coord_sf(xlim = c(-185, -60)) +
  theme(legend.position = "bottom")

```

  

Spatial Graph 3: Highlighting states with specific policies on spatial graph of total # of female incarcerated to per 100,000 in each state

```{r figure 7, echo = T, message=F, warning=F}

#Highlight specific states 

strict_prositution_states <- states_incar_fpc %>% 
  filter(NAME %in% c("Oklahoma", "Texas", "Utah", "Pennsylvania", "Nevada", "Kentucky", 
                     "West Virginia", "Vermont", "Louisiana", "Idaho", "Indiana", 
                     "Illinois"))

head(strict_prositution_states) #new data set for states with stricter prostitution laws

#Number of Females Incarcerated Per 100,000 People in Each State, Highlighting States 
#with Strict Prostitution Laws in Green

ggplot(states_incar_fpc)+
  geom_sf(aes(fill = `Number of Females Incarcerated Per 100,000 People`))+
  scale_fill_viridis(option = "magma", direction = -1)+
  labs(title = "Number of Females Incarcerated Per 100,000 People in Each State", 
       subtitle = "States with Stricter Punishment on Prostitution are Highlighted in 
Green Borders")+
  geom_sf(data = strict_prositution_states, 
          fill = NA, 
          color = "green",
          size = 0.5) +
  coord_sf(xlim = c(-185, -60)) +
  theme(legend.position = "bottom")

```

  
T-Test: Highlighting states with specific policies on spatial graph of total # of female incarcerated to male incarcerated per state

If the p-value is less than 0.05 (or 5%), the result is considered statistically significant. 

```{r t-test, echo = T, message=F, warning=F}

#Add row for states with stricter punishments on prostitution vs not
strictness_states_incar_fpc <- states_incar_fpc %>% 
  mutate(Strictness=case_match(NAME,
                             #Strict
                             c("Oklahoma", "Texas", "Utah", "Pennsylvania", "Nevada", 
                               "Kentucky", "West Virginia", "Vermont", "Louisiana", 
                               "Idaho", "Indiana", "Illinois") ~ "Strict",
                             c("South Carolina", "Maine", "New Hampshire", "Massachusetts", 
                               "Connecticut", "Rhode Island", "New York", "California", 
                               "New Jersey", "North Dakota", "South Dakota", "Nebraska", 
                               "Kansas", "Minnesota", "Iowa", "Missouri",  
                               "Wisconsin", "Michigan", "Ohio", "Washington", "Oregon", 
                               "Wyoming", "Montana", "Colorado", "Arizona", "New Mexico", 
                               "Arkansas", "Mississippi", "Tennessee", "Alabama", "Georgia", 
                               "Florida", "North Carolina", "Maryland", "Delaware", 
                               "Pennsylvania", "Alaska", "Hawaii", "Virginia") ~ "Not Strict")) %>%
  group_by(Strictness) %>% mutate(sum = sum(Total, na.rm = TRUE))

head(strictness_states_incar_fpc) #shows new column that organizes states as strict vs not strict

#T-test of strict vs not strict population Number of Females Incarcerated per 100,000 in 
#State Population
t.test(`Number of Females Incarcerated Per 100,000 People` ~ Strictness, 
       data = strictness_states_incar_fpc)


```

