---
title: "Final Project Proposal"
author: "Your Name Here! (group members here, optional)"
date: "Date goes Here!"
output: pdf_document
---


#check to make sure say per 100,000 everywhere instead of per capita 

## Final Project Proposal
Proposals should answer the following questions:

1. What is your research question? Why does it matter? How does it relate to society?

2. What data will you use to answer the question? Do you have access to these data?

3. What methods will you use to answer your research question?

4. What kind of product will you create? How do you want others to read or engage with your project?


Download all necessary libraries
```{r libraries, echo = T, message=F, warning=F}
#Libraries
library(RCurl)
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(sf)
library(tidycensus)
library(tigris)
library(viridis)

```


Download data 
```{r data, echo = T, message=F, warning=F}

url <- getURL("https://raw.githubusercontent.com/avaelisa/soc10finalproject-alex-christine-ava-luci/refs/heads/main/soc10-dataset")

# download incarceration data (from github)
data <- read_csv(url)

#View(data)

```

Data processing part 1: fix names, prep per 100,000, and spatial data 

```{r data processing 1, echo = T, message=F, warning=F}

#Mutate to fix state names and remove U.S. total, Federal, and All States
data %<>%
  mutate(Location=case_match(Location,
                             "Alaska/b"~"Alaska",
                             "Connecticut/b"~"Connecticut",
                             "Delaware/b"~"Delaware",
                             "Hawaii/b"~"Hawaii",
                             "Maryland/c"~"Maryland",
                             "Missouri/d"~"Missouri",
                             "Oklahoma/c"~"Oklahoma",
                             "Rhode Island/b"~"Rhode Island",
                             "Vermont/b"~"Vermont",
                             .default=Location)) %>%
                            drop_na(Total) %>% #get rid of district of columbia and puerto rico because there is no data
  filter(!Location %in% c("U.S. total", "Federal", "All States"))

#census API
census_api_key("857ead205d4b6101ae723a246e9b6f95b1b827d4")

#add per capita with tidycensus
states <- get_acs(
  geography = "state",
  variables = "B01003_001",
  geometry = TRUE,
  year = 2023) %>%
  filter(!NAME %in% c("District of Columbia", "Puerto Rico")) #remove these because no data as seen above 

#combine data: USA map + incarceration rate
states_incar <- states %>%
  left_join(data, by = c("NAME" = "Location"))

#add female per one hundered thousand people to the spatial data
states_incar_fpc <- states_incar %>%
  mutate("Female Incarceration Ratio Per 100,000" = Female/estimate*100000)

#create column of incarceration per 1000,000 males 
#add male per one hundered thousand people to the spatial data
states_incar_fmpc <- states_incar_fpc %>%
  mutate("Male Incarceration Ratio Per 100,000" = Male/estimate*100000)
states_incar_fmpc <- states_incar_fpc %>%
  mutate("Male Incarceration Ratio Per 100,000" = Male/estimate*100000)

```


Bar Chart 1A: Total Population Incarcerated by State
Goal: Compare incarceration rates for all states

Using coord_flip to make bar chart horizontal. Reorder in ggplot will rank bars from smallest to biggest so it is easier to see trends and which states have highest vs lowest incarceration rates. This will make data easier to analyze. 

```{r figure 1A, echo = T, message=F, warning=F}

state_total_graph <- ggplot(data, aes(x = reorder(Location, Total), y = Total))+ #reorder 
  geom_bar(position = "dodge", stat = "identity", 
           aes(fill = Location))+
           labs(y = "Total", title = "Total Population Incarcerated by State")+
             coord_flip() 


#See visualization 
state_total_graph 

```


Bar Chart 1B: Total Population Incarcerated by State Per 100,000 Capita
Goal: Compare incarceration rates for all states

```{r figure 1B, echo = T, message=F, warning=F}

state_total_graph <- ggplot(states_incar_fpc, aes(x = reorder(NAME, `Female Incarceration Ratio Per 100,000`), y = `Female Incarceration Ratio Per 100,000`))+ #reorder 
  geom_bar(position = "dodge", stat = "identity", 
           aes(fill = NAME))+
           labs(x = "", y = "Total", title = "Total Population Incarcerated by State")+
             coord_flip() 


#See visualization 
state_total_graph 

```



Bar Chart 2A: Total Incarceration Rate with States Divided by Region 

The states were divided into regions through case_match.

```{r figure 2, echo = T, message=F, warning=F}

region_data <- data %>% 
  mutate(Region=case_match(Location,
                             #Northeast
                             c("Maine", "Vermont", "New Hampshire", "Massachusetts", "Connecticut", "Rhode Island", "New York", "Pennsylvania", "New Jersey", "Virginia") ~ "Northeast",
                             #Midwest
                             c("North Dakota", "South Dakota", "Nebraska", "Kansas", "Minnesota", "Iowa", "Missouri", "Illinois", "Wisconsin", "Michigan", "Ohio", "Indiana") ~ "Midwest",
                             #West
                             c("Washington", "Oregon", "California", "Nevada", "Idaho", "Wyoming", "Montana", "Colorado", "Utah", "Arizona", "New Mexico") ~ "West",
                             #South
                             c("Texas", "Oklahoma", "Arkansas", "Louisiana", "Mississippi", "Kentucky", "Tennessee", "Alabama", "Georgia", "West Virginia", "Florida", "South Carolina", "North Carolina", "Maryland", "Delaware") ~ "South",
                             .default = "Other")) %>%
                            drop_na(Total) %>%
  group_by(Region) %>% mutate(sum = sum(Total, na.rm = TRUE))

#View(region_data)

#Plot
region_total_graph <- 
  ggplot(region_data, aes(x = reorder(Region, Total), y = sum)) +
  geom_bar(aes(fill = Region), position = "dodge", stat = "identity")+
  labs(x = "Region", y = "Total", title = "Total Population Incarcerated by Region")+
  theme_linedraw()+
  coord_flip()

#Print Plot
region_total_graph

```



Bar Chart 2A: Total Incarceration Rate with States Divided by Region Per 100,000 Population

The states were divided into regions through case_match.

```{r figure 2B, echo = T, message=F, warning=F}


region_data_pc <- states_incar_fmpc %>% 
  mutate(Region=case_match(NAME,
                             #Northeast
                             c("Maine", "Vermont", "New Hampshire", "Massachusetts", "Connecticut", "Rhode Island", "New York", "Pennsylvania", "New Jersey", "Virginia") ~ "Northeast",
                             #Midwest
                             c("North Dakota", "South Dakota", "Nebraska", "Kansas", "Minnesota", "Iowa", "Missouri", "Illinois", "Wisconsin", "Michigan", "Ohio", "Indiana") ~ "Midwest",
                             #West
                             c("Washington", "Oregon", "California", "Nevada", "Idaho", "Wyoming", "Montana", "Colorado", "Utah", "Arizona", "New Mexico") ~ "West",
                             #South
                             c("Texas", "Oklahoma", "Arkansas", "Louisiana", "Mississippi", "Kentucky", "Tennessee", "Alabama", "Georgia", "West Virginia", "Florida", "South Carolina", "North Carolina", "Maryland", "Delaware") ~ "South",
                             .default = "Other")) %>%
                            drop_na(Total) %>%
  group_by(Region) %>% mutate(sum = sum(`Female Incarceration Ratio Per 100,000`, na.rm = TRUE))

#View(region_data_pc)


#Plot
region_total_graph <- 
  ggplot(region_data_pc, aes(x = reorder(Region, estimate), y = sum)) +
  geom_bar(aes(fill = Region), position = "dodge", stat = "identity")+
  labs(x = "Region", y = "Total", title = "Total Population Incarcerated by Region Per 100,000 Capita")+
  theme_linedraw()+
  coord_flip()

#Print Plot
region_total_graph

```

Bar Chart 3A: Total Incarceration Rate with States Divided by State by Gender

```{r figure 3A, echo = T, message=F, warning=F}

#pivot longer
state_data <- data %>%
  pivot_longer(cols = c(Male, Female),
               names_to = "Gender",
               values_to = "Value")
 
#plot              
state_total_stacked <- ggplot(state_data, aes(x = Location, y = Value))+
  geom_bar(position = "stack", stat = "identity",
           aes(fill = Gender))+
  theme_linedraw()+
  labs(x = "State Separated by Gender", title = "Total Population Incarcerated by State and Gender")+
  coord_flip() 
  
#print plot 
state_total_stacked

```

Bar Chart 3B: Total Incarceration Rate with States Divided by State by Gender  Per 100,000 People

```{r figure 3B, echo = T, message=F, warning=F}

#pivot longer
gendered_states_incar_fmpc <- states_incar_fmpc %>%
  pivot_longer(cols = c(`Male Incarceration Ratio Per 100,000`, `Female Incarceration Ratio Per 100,000`),
               names_to = "Gender",
               values_to = "Value")
 
#plot              
state_total_stacked <- ggplot(gendered_states_incar_fmpc, aes(x = NAME, y = Value))+
  geom_bar(position = "stack", stat = "identity",
           aes(fill = Gender))+
  theme_linedraw()+
  labs(x = "State Separated by Gender", title = "Total Population Incarcerated Per 100,000 by State and Gender")+
  coord_flip() 
  
#print plot 
state_total_stacked

```




Bar Chart 4A: Total Incarceration Rate with States Divided by Region by Gender

```{r figure 4A, echo = T, message=F, warning= F}

#pivot longer
region_data2 <- region_data %>%
  pivot_longer(cols = c(Male, Female),
               names_to = "Gender",
               values_to = "Value")
              
    
#plot 
gender_by_region <- ggplot(region_data2, aes(x = Region, y = Value))+
  geom_bar(position = "stack", stat = "identity",
           aes(fill = Gender))+
  theme_linedraw()+
  labs(x = "Region Separated by Gender",  title = "Total Population Incarcerated by Region and Gender")+
  coord_flip() 
 
#print plot  
gender_by_region

```
Bar Chart 4B: Total Incarceration Rate with States Divided by Region by Gender

```{r figure 4B, echo = T, message=F, warning= F}

#pivot longer
region_data_pc2 <- region_data_pc %>%
  pivot_longer(cols = c(`Male Incarceration Ratio Per 100,000`, `Female Incarceration Ratio Per 100,000`),
               names_to = "Gender",
               values_to = "Value")
              
    
#plot 
gender_by_region_pc_graph <- ggplot(region_data_pc2, aes(x = Region, y = Value))+
  geom_bar(position = "stack", stat = "identity",
           aes(fill = Gender))+
  theme_linedraw()+
  labs(x = "Region Separated by Gender",  title = "Total Population Incarcerated by Region and Gender Per 100,000")+
  coord_flip() 
 
#print plot  
gender_by_region_pc_graph

```



Set up spatial data - I thinK WE CAN DEKETE ThIS

```{r set up spatial data, echo = T, message=F, warning=F}


#get map of states in the US and pull population data #i think can delete this
#states <- get_acs(
  #geography = "state",
  #variables = "B01003_001",
  #geometry = TRUE,
 # year = 2023)

#combine data: USA map + incarceration rate #i think can delete this 
#states_incar <- states %>%
  #left_join(data, by = c("NAME" = "Location"))


```


Spatial Graph 1: Total # of women incarcerated per 100,000 people

```{r figure 5, echo = T, message=F, warning=F}

#add female per one hundered thousand people to the spatial data i think can delete this 
#states_incar_fpc <- states_incar %>%
  #mutate("Female Incarceration Ratio Per 100,000" = Female/estimate*100000)
  
#plot combined data - Female Incarceration Ratio Per 100,000 in Each State
ggplot(states_incar_fpc)+
  geom_sf(aes(fill = `Female Incarceration Ratio Per 100,000`))+
  scale_fill_viridis(option = "magma", direction = -1)+
  labs(title = "Female Incarceration Ratio Per 100,000 in Each State") +
  coord_sf(xlim = c(-185, -60))

```


Spatial Graph 2: Ratio of total # of female incarcerated to male incarcerated per State

```{r figure 6, echo = T, message=F, warning=F}

#add female per male to spatial data 

states_incar_fpm <- states_incar_fmpc %>%
  mutate("Female per 100,000 population to Male per 100,000 Ratio" = `Female Incarceration Ratio Per 100,000`/ `Male Incarceration Ratio Per 100,000`)

#plot combined data - Ratio of Female to Male
ggplot(states_incar_fpm)+
  geom_sf(aes(fill = `Female per 100,000 population to Male per 100,000 Ratio`))+
  scale_fill_viridis(option = "magma", direction = -1)+
  labs(title = "Total Female to Total Male Incarceration Ratio in Each State")  +
  coord_sf(xlim = c(-185, -60))

```

  

Spatial Graph 3: Highlighting states with specific policies on spatial graph of total # of female incarcerated to per capita arrests

```{r figure 7, echo = T, message=F, warning=F}

#highlight specific states 

strict_prositution_states <- states_incar_fpc %>% 
  filter(NAME %in% c("South Carolina", "Oklahoma", "Texas", "Utah", "Pennsylvaia", "Nevada", "Kentucky", "West Virginia", "Vermont", "Louisiana", "Idaho", "Indiana", "Illinois"))

#Female Incarceration Ratio Per 100,000 in Each State Highlighting those with Strict Laws in White
ggplot(states_incar_fpc)+
  geom_sf(aes(fill = `Female Incarceration Ratio Per 100,000`))+
  scale_fill_viridis(option = "magma", direction = -1)+
  labs(title = "Female Incarceration Ratio Per 100,000 in Each State", 
       subtitle = "States with Stricter Stances on Position are Highlighted in White Borders")+
  geom_sf(data = strict_prositution_states, 
          fill = NA, 
          color = "green",
          size = 8) +
  coord_sf(xlim = c(-185, -60))

```

  
T-Test: Highlighting states with specific policies on spatial graph of total # of female incarcerated to male incarcerated per state

p-value is typically less than 0.05 (or 5%), the result is considered statistically significant. used default two tailed t-test. 

```{r t-test, echo = T, message=F, warning=F}

#add row for those with stricter take on prostitution vs not
strictness_states_incar_fpc <- states_incar_fpc %>% 
  mutate(Strictness=case_match(NAME,
                             #Strict
                             c("South Carolina", "Oklahoma", "Texas", "Utah", "Pennsylvaia", "Nevada", "Kentucky", "West Virginia", "Vermont", "Louisiana", "Idaho", "Indiana", "Illinois") ~ "Strict",
                             c("Maine", "New Hampshire", "Massachusetts", "Connecticut", "Rhode Island", "New York", "New Jersey", "North Dakota", "South Dakota", "Nebraska", "Kansas", "Minnesota", "Iowa", "Missouri", "Illinois", "Wisconsin", "Michigan", "Ohio", "Washington", "Oregon", "California", "Wyoming", "Montana", "Colorado", "Arizona", "New Mexico", "Arkansas", "Mississippi", "Tennessee", "Alabama", "Georgia", "Florida", "North Carolina", "Maryland", "Delaware", "Pennsylvania", "Alaska", "Hawaii", "Virginia") ~ "Not Strict", 
         .default = "Other")) %>%
  group_by(Strictness) %>% mutate(sum = sum(Total, na.rm = TRUE))

#View(strictness_states_incar_fpc)

#ttest 
t.test(`Female Incarceration Ratio Per 100,000` ~ Strictness, data = strictness_states_incar_fpc)


```

